Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-11-01 manuscript/manuscript.Rmd
Edits from JW

Key words: Variant interpretation, Clinical Genomics, Allele frequency, Cardiomyopathy, Inherited Cardiovascular conditions, ExAC

@@

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency filtering in Mendelian variant discovery efforts.

In practice, there exists considerable ambiguity around what allele frequency should be considered "too common", with the lenient values of 1% and 0.1% often invoked as conservative frequency cutoffs for recessive and dominant diseases respectively `r citep("10.1038/nrg3031")`. Population genetics, however, dictates that severe disease-causing variants must be much rarer than these cutoffs, except in cases of bottlenecked populations, balancing selection, or other special circumstances `r citep(c("10.1093/molbev/msp190","10.1073/pnas.1322563111"))`.

It is intuitive that when assessing a variant for a causative role in a dominant Mendelian disease, the frequency of a variant in a reference sample, not selected for the condition, should not exceed the prevalence of the condition `r citep(c("10.1016/j.ajhg.2016.03.024","10.1126/scitranslmed.aad5169"))`. This rule must, however, be refined to account for different inheritance modes, genetic and allelic heterogeneity, and reduced penetrance. In addition, for rare variants, estimation of true population allele frequency is clouded by considerable sampling variance, even in the largest samples currently available. These limitations have encouraged the adoption of very lenient approaches when filtering variants by allele frequency `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that allele frequency cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed allele frequency filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying our framework.

@@

For a penetrant dominant Mendelian allele to be disease causing, it cannot be present in the general population more frequently that the disease it causes. Furthermore, if the disease is genetically heterogeneous, it must not be more frequent than the proportion of cases attributable to that gene, or indeed to any single variant. We can therefore define the maximum credible population allele frequency (for a pathogenic allele) as:

@@

For recessive conditions, the maximum AF is defined as:
*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant. Full derivation can be found in **Supplementary Methods**.

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. For cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.

We do not know the true population allele frequency of any variant, having only an observed allele frequency in a finite population sample. Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants. For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000. `r citep("10.1038/nature19057")`

If we turn the problem around, and begin instead from allele frequency, specifying a maximum _true_ allele frequency value we are willing to consider in the population (using the equation above), then we can estimate the probability distribution for allele counts in a given sample size. This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given population frequency.

The maximum tolerated allele count (AC) was therefore computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN). In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.

@@

In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population (more explanation in **Supplementary Methods**).

@@

To simulate Mendelian variant discovery, we randomly selected 100 individuals from each of five major continental populations and filtered their exomes against filtering allele frequencies derived from the remaining 60,206 ExAC individuals. The subset of individuals was the same as that previously reported `r citep("10.1038/nature19057")`.
Predicted protein-altering variants are defined as missense and equivalent (including in-frame indels, start lost, stop lost, and mature miRNA-altering), and protein-truncating variants (nonsense, essential splice site, and frameshift).

@@

Pathogenic and non-conflicted variants from XML and TXT releases of ClinVar (July 9, 2015 release) were extracted using a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

@@

#### *Application and validation in hypertrophic cardiomyopathy*
#### *Defining maximum credible population allele frequency*

We illustrate our general approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can make the assumption that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries). This allows us to define the maximum allelic contribution of any single variant to the disorder. In these large case series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population allele frequency for this allele, assuming 50% penetrance as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1). 
#### *Controlling for sample variation*
To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population allele frequency of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population allele frequency (**Methods**). For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population allele frequency and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.

#### *Assessing the acuracy of our approach*
For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).

@@

To assess the HCM thresholds empirically, we explored the ExAC allele frequency spectrum of `r myVariantsN` distinct autosomal variants identified in 6179 recently published HCM cases referred for diagnostic sequencing, and individually assessed and reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1). This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are unlikely to be truly causative for HCM.

The above analysis applied a single global allele count limit of 9 for HCM, however, as allele frequencies differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`. For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations. We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest allele frequency observed in any major continental population (see **Methods** and **Suplementary Note 3**).

To further validate this approach, we examined all 601 variants identified in ClinVar `r citep("10.1093/nar/gkt1113")` as "Pathogenic" or "Likely Pathogenic" and non-conflicted for HCM. 558 (93%) were sufficiently rare when assessed as described. 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated. 42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and would be classified by the ACMG framework as VUS at most. The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population. This fell outside the 95% confidence interval for an underlying population frequency <4x10^-5^, but within the 99% confidence threshold: a single outlier due to stochastic variation is unsurprising given that these nominal probabilities are not corrected for multiple testing across 601 variants. In light of our updated assessment, 20 variants were reclassified as Benign/Likely Benign and 22 as VUS according to the American College for Medical Genetics and Genomics (ACMG) guidelines for variant interpretation `r citep("10.1038/gim.2015.30")` (Table S1).

@@

In the absence of large case series, we must estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.

Where disease-specific variant databases exist, we can use these to help estimate the maximum allelic contribution in lieu of individual case series. For example, Marfan syndrome is a rare connective tissue disorder caused by variants in the _FBN1_ gene. The UMD-FBN1 database `r citep("10.1002/humu.10249")` contains 3077 variants in _FBN1_ from 280 references (last updated 28/08/14). The most common variant is in 30/3006 records (1.00%; 95CI 0.53-1.46%), which likely overestimates its contribution to disease if related individuals are not systematically excluded. Taking the upper bound of this frequency as our maximum allelic contribution, we derive a maximum tolerated allele count of 2 (Table 2). None of the five most common variants in the database are present in ExAC.

Where no mutation database exists, we can use what is known about similar disorders to estimate the maximum allelic contribution. For the cardiac conditions with large cases series in Table 1, the maximum proportion of cases attributable to any one variant is 6.7% (95CI 4.1-9.2%; PKP2:c.2146-1G>C found in 24/361 ARVC cases `r citep("10.1038/gim.2016.90")`). We therefore take the upper bound of this confidence interval (rounded up to 0.1) as an estimate of the maximum allelic contribution for other genetically heterogeneous cardiac conditions, unless we can find disease-specific evidence to alter it. For Noonan syndrome and Catecholaminergic Polymorphic Ventricular Tachycardia (CPVT - an inherited cardiac arrhythmia syndrome) with prevalences of 1 in 1000 `r citep("10.1016/S0140-6736(12)61023-X")` and 1 in 10,000 `r citep("10.1038/ejhg.2013.55")` respectively, this translates to maximum population frequencies of 5x10^-5^ and 5x10^-6^ and maximum tolerated ExAC allele counts of 10 and 2 (Table 2).

@@

Intuitively, if one penetrant recessive variant were to be responsible for all PCD cases, it could have a maximum population frequency of $\sqrt(1/10000)$. We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution (see *Methods*). Across previously published cohorts of PCD cases `r citep(c("10.1164/rccm.200603-370OC","10.1164/rccm.200601-084OC","10.1038/ng.2277"))`, *DNAI1* IVS1+2\_3insT was the most common variant with a total of 17/358 alleles (4.7% 95CI 2.5-7.0%). Given that ~9% of all patients with PCD have disease-causing variants in _DNAI1_ and the IVS1+2\_3insT variant is estimated to account for ~57% of variant alleles in _DNAI1_ `r citep("10.1164/rccm.200603-370OC")`, we can take these values as estimates of the maximum genetic and allelic contribution for PCD, yielding a maximum expected population AF of $\sqrt(1/10000) \times 0.57 \times \sqrt0.09 \times 1/\sqrt0.5 = 2.42 \times10$^-3^ This translates to a maximum tolerated ExAC AC of 322. *DNAI1* IVS1+2\_3insT is itself present at 56/121108 ExAC alleles (45/66636 non-Finnish European alleles). A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and have reclassified it as VUS (see **Supplementary Methods**).

@@

For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering allele frequency≥the maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate. The filtering allele frequency is not desease speific and hence can be applied to any disease of interest by comparing with a user-defined disease-specific maximum credible allele frequency (Figure 3). This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

To assess the efficiency of our approach, we calculated the filtering allele frequency on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**). Filtering at allele frequencies lower than 0.1% substantially reduces the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 to 63 at cutoffs of 0.1% and 0.0001% respectively (Figure 2a). Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the allele frequency spectrum, and computed disease odds ratios for different frequency bins. The odds ratio for disease-association increases markedly at very low allele frequencies (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.

@@

To assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset. This is defined such that if the filtering allele frequency of a variant is at or above the user-defined "maximum credible population allele frequency" for the disease in question, then that variant is not a credible candidate (in other words, for any population allele frequency below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05). Once a user has determined their "maximum credible population allele frequency", they may remove from consideration ExAC variants for which the filtering allele frequency is greater than or equal to than the chosen value.

@@

We recognise several limitations of our approach. First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. Unfortunately, there is no substitute for understanding the genetic architecture of the disease of interest, but when considering a specific variant that has been observed at low frequency in a given reference population, our framework and accompanying web tool provide a transparent approach to define the range of disease architectures that are compatible with the observed data. For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.

>>>>

Key words: Variant interpretation, Clinical Genomics, Allele frequency, Cardiomyopathy, Inherited Cardiovascular Conditions, ExAC

@@

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency (AF) filtering in Mendelian variant discovery efforts.

In practice, there exists considerable ambiguity around what AF should be considered "too common", with the lenient values of 1% and 0.1% often invoked as conservative frequency cutoffs for recessive and dominant diseases respectively `r citep("10.1038/nrg3031")`. Population genetics, however, dictates that severe disease-causing variants must be much rarer than these cutoffs, except in cases of bottlenecked populations, balancing selection, or other special circumstances `r citep(c("10.1093/molbev/msp190","10.1073/pnas.1322563111"))`.

It is intuitive that when assessing a variant for a causative role in a dominant Mendelian disease, the frequency of a variant in a reference sample, not selected for the condition, should not exceed the prevalence of the condition `r citep(c("10.1016/j.ajhg.2016.03.024","10.1126/scitranslmed.aad5169"))`. This rule must, however, be refined to account for different inheritance modes, genetic and allelic heterogeneity, and reduced penetrance. In addition, for rare variants, estimation of true population AF is clouded by considerable sampling variance, even in the largest samples currently available. These limitations have encouraged the adoption of very lenient approaches when filtering variants by AF `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that AF cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed AF filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying this framework.

@@

We define a two-stage approach to determine whether a variant observed in a reference sample is too common to cause a given disease.  First, we define a maximum population AF that we believe is credible for a pathogenic variant, given the genetic architecture of the disease in question. Second, we determine whether the observed allele count in our reference sample is consistent with a variant having this frequency in the population from which the sample was drawn.

For a penetrant dominant Mendelian allele to be disease causing, it cannot be present in the general population more frequently that the disease it causes. Furthermore, if the disease is genetically heterogeneous, it must not be more frequent than the proportion of cases attributable to that gene, or indeed to any single variant. We can therefore define the maximum credible population AF (for a pathogenic allele) as:

@@

For recessive conditions, the maximum AF is defined as:
*maximum credible population AF* = $\sqrt($*prevalence*$) \times$ *maximum allelic contribution* $\times \sqrt($*maximum genetic contribution*$) \times 1/\sqrt($*penetrance*$)$

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant (see **Supplementary Methods** for full derivation).

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.

We do not know the true population AF of any variant, having only an observed AF in a finite population sample. Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants. For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000. `r citep("10.1038/nature19057")`

To address this, we begin by specifying a maximum _true_ AF value we are willing to consider in the population (using the equation above), from which we can estimate the probability distribution for allele counts in a given sample size. This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given underlying population frequency. For example, a variant with a true population AF of 0.0001 would be expected to occur in a sample of 100,000 alleles ≤15 times with a probability of 0.95.

We therefore computed a maximum tolerated allele count (AC) as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the specified maximum credible AF, given the sample size (observed allele number, AN).

@@

We can reverse this process to determine the maximum true population AF that is consistent with a particular observed sample AC, and we applied this to the ExAC dataset (version 0.3.1). In order to pre-compute `af_filter` values for all variants in ExAC, we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population (more explanation in **Supplementary Methods**).

@@

To simulate Mendelian variant discovery, we randomly selected 100 individuals from each of five major continental populations and filtered their exomes against filtering AFs derived from the remaining 60,206 ExAC individuals. The subset of individuals was the same as that previously reported `r citep("10.1038/nature19057")`.
Predicted protein-altering variants are defined as missense and equivalent (including in-frame indels, start lost, stop lost, and mature miRNA-altering), and protein-truncating variants (nonsense, essential splice site, and frameshift).

@@

Pathogenic and non-conflicted variants were extracted from ClinVar (July 9, 2015 release) as described previously `r citep("10.1038/nature19057")`. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

@@

#### *Application and validation in hypertrophic cardiomyopathy*
#### *Defining maximum credible population AF*

We illustrate our generalisable approach using the dominant cardiac disorder hypertrophic cardiomyopathy (HCM), which has an estimated prevalence of 1 in 500 in the general population `r citep("10.1038/gim.2014.205")`. As there have been previous large-scale genetic studies of HCM, with series of up to 6,179 individuals `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, we can assume that no newly identified variant will be more frequent in cases that those identified to date (at least for well-studied ancestries), allowing us to define the maximum contribution of any single variant to the disorder. In these series, the largest proportion of cases is attributable to the missense variant _MYBPC3_ c.1504C>T (p.Arg502Trp), found in 104/6179 HCM cases (1.7%; 95%CI 1.4-2.0%) `r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. We therefore take the upper bound of this proportion (0.02) as an estimate of the maximum allelic contribution in HCM (Table 1). Our maximum expected population AF for this allele, assuming penetrance 0.5 as previously reported `r citep("10.1161/CIRCRESAHA.109.216291")`, is 1/500 x 1/2 (dividing prevalence per individual by the number of chromosomes per individual) x 0.02 x 1/0.5 = 4.0x10^-5^, which we take as the maximum credible population AF for any causative variant for HCM (Table 1).

#### *Controlling for sample variation*

To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population AF of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population AF (**Methods**). For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population AF and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.

#### *Assessing the acuracy of our approach*

For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).

@@

To assess the HCM thresholds empirically, we explored the ExAC AF spectrum of `r myVariantsN` distinct autosomal variants, identified in 6179 published HCM cases referred for diagnostic sequencing, and individually assessed and clinically reported according to international guidelines`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`. 
`r myPassPath`/`r myAllPath` (`r signif(100*myPassPath/myAllPath,3)`%) variants reported as ‘Pathogenic’ or ‘Likely Pathogenic’ fell below our threshold (Figure 1), including all variants with a clear excess in cases. The `r myFailPath` variants historically classified as ‘Likely Pathogenic’, but prevalent in ExAC in this analysis, were reassessed using contemporary ACMG criteria: there was no strong evidence in support of pathogenicity, and they were reclassified in light of these findings (Table S1). This analysis identifies `r myFailVus`/`r myAllVus` (`r signif(100*myFailVus/myAllVus,3)`%) VUS that are very unlikely to be causative for HCM.

The above analysis applied a single global allele count limit of 9 for HCM, however, as AFs differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`. For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations. We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest AF observed in any major continental population (see **Methods** and **Suplementary Note 3**).

To further validate this approach, we examined all 601 variants identified in ClinVar `r citep("10.1093/nar/gkt1113")` as "Pathogenic" or "Likely Pathogenic" and non-conflicted for HCM. 558 (93%) were sufficiently rare when assessed as described. 43 variants were insufficiently rare in at least one ExAC population, and were therefore re-curated. 42 of these had no segregation or functional data sufficient to demonstrate pathogenicity in the heterozygous state, and would be classified by the contemporary ACMG framework as VUS at most. The remaining variant (MYBPC3:c.3330+5G>C) had convincing evidence of pathogenicity, though with uncertain penetrance (see **Supplementary Methods**), and was observed twice in the African/African American ExAC population. This fell outside the 95% confidence interval for an underlying population frequency <4x10^-5^, but within the 99% confidence threshold: a single outlier due to stochastic variation is unsurprising given that these nominal probabilities are not corrected for multiple testing across 601 variants. In light of our updated assessment, 20 variants were reclassified as Benign/Likely Benign and 22 as VUS according to the American College for Medical Genetics and Genomics (ACMG) guidelines for variant interpretation `r citep("10.1038/gim.2015.30")` (Table S1).

@@

This framework relies on estimation of the genetic architecture of a condition, which may not be well described. For diseases where large case series are absent, we can estimate the genetic architecture parameters by extrapolating from similar disorders and/or variant databases.

Where disease-specific variant databases exist, we can use these to estimate the maximum allelic contribution. For example, Marfan syndrome is a rare connective tissue disorder caused by variants in the _FBN1_ gene. The UMD-FBN1 database `r citep("10.1002/humu.10249")` contains 3077 variants in _FBN1_ from 280 references (last updated 28/08/14). The most common variant is in 30/3006 records (1.00%; 95CI 0.53-1.46%), which likely overestimates its contribution to disease if related individuals are not systematically excluded. Taking the upper bound of this frequency as our maximum allelic contribution, we derive a maximum tolerated allele count of 2 (Table 2). None of the five most common variants in the database are present in ExAC.

Where no mutation database exists, we can use what is known about similar disorders to estimate the maximum allelic contribution. For the better-characterised cardiac conditions in Table 1, the maximum proportion of cases attributable to any one variant is 6.7% (95CI 4.1-9.2%; PKP2:c.2146-1G>C found in 24/361 ARVC cases `r citep("10.1038/gim.2016.90")`). We therefore propose the upper bound of this confidence interval (rounded up to 0.1) as a resonable estimate of the maximum allelic contribution for other genetically heterogeneous cardiac conditions, unless there is disease-specific evidence to alter it. For Noonan syndrome and Catecholaminergic Polymorphic Ventricular Tachycardia (CPVT - an inherited cardiac arrhythmia syndrome) with prevalences of 1 in 1000 `r citep("10.1016/S0140-6736(12)61023-X")` and 1 in 10,000 `r citep("10.1038/ejhg.2013.55")` respectively, this translates to maximum population frequencies of 5x10^-5^ and 5x10^-6^ and maximum tolerated ExAC allele counts of 10 and 2 (Table 2).

@@

Intuitively, if one penetrant recessive variant were to be responsible for all PCD cases, it could have a maximum population frequency of $\sqrt(1/10000)$. We can refine our evaluation of PCD by estimating the maximum genetic and allelic contribution (see *Methods*). Across previously published cohorts of PCD cases `r citep(c("10.1164/rccm.200603-370OC","10.1164/rccm.200601-084OC","10.1038/ng.2277"))`, *DNAI1* IVS1+2\_3insT was the most common variant with a total of 17/358 alleles (4.7% 95CI 2.5-7.0%). Given that ~9% of all patients with PCD have disease-causing variants in _DNAI1_ and the IVS1+2\_3insT variant is estimated to account for ~57% of variant alleles in _DNAI1_ `r citep("10.1164/rccm.200603-370OC")`, we can take these values as estimates of the maximum genetic and allelic contribution for PCD, yielding a maximum expected population AF of $\sqrt(1/10000) \times 0.57 \times \sqrt0.09 \times 1/\sqrt0.5 = 2.42 \times10$^-3^ This translates to a maximum tolerated ExAC AC of 322. *DNAI1* IVS1+2\_3insT is itself present at 56/121108 ExAC alleles (45/66636 non-Finnish European alleles). A single variant reported to cause PCD in ClinVar occurs in ExAC with AC > 332 (_NME8_ NM\_016616.4:c.271-27C>T; AC=2306/120984): our model therefore indicates that this variant frequency is too common to be disease-causing, and consistent with this we note that it meets none of the current ACMG criteria for assertions of pathogenicity, and would reclassify it as VUS (see **Supplementary Methods**).

@@

For each ExAC variant, we defined a "filtering AF" that represents the threshold disease-specific "maximum credible AF" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering AF ≥ the maximum credible AF for the disease under consideration should be filtered, while a variant with a filtering AF below the maximum credible remains a candidate. This filtering AF is not disease specific: it can be applied to any disease of interest by comparing with a user-defined disease-specific maximum credible AF (Figure 3). This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

To assess the efficiency of our approach, we calculated the filtering AF on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**). Filtering at AFs lower than 0.1% substantially reduces the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 to 63 at cutoffs of 0.1% and 0.0001% respectively (Figure 2a). Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the AF spectrum, and computed disease odds ratios for different frequency bins. The odds ratio for disease-association increases markedly at very low AFs (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.

@@

To assist others in applying our framework, we have precomputed a 'filtering AF' for all variants across the ExAC dataset. This is defined such that if the filtering AF of a variant is at or above the user-defined "maximum credible population AF" for the disease in question, then that variant is not a credible candidate (in other words, for any population AF below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05). Once a user has determined their "maximum credible population AF", they may remove from consideration ExAC variants for which the filtering AF is greater than or equal to than the chosen value.

@@

We recognise several limitations of our approach. First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. While incomplete knowledge of the genetic architecture of a disease of interest will limit this or any approach to evaluate a specific variant that has been observed at low frequency in a reference population, our framework and accompanying web tool do at least transparently define the range of disease architectures that are compatible with the observed data. For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.

Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants. Although we would argue that variants with low penetrance have questionable diagnostic utility, our calculator app allows a user to define a range of compatible penetrance for a given AF (see **Supplementary Methods**), and implements methods to estimate variant penetrance from prevalence data in case and control cohorts as previously described `r citep("10.1126/scitranslmed.aad5169")`.

Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts. Where this is possible, the maximum credible population AF should be derived based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence.
